---
title: Anomaly Detection Ensembles
author: Sevvandi Kandanaarachchi
date: '2022-03-16'
slug: []
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2022-03-16T15:50:32+11:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, message=FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(here)
library(R.matlab)

# Unitize each column of X
unitize <- function(X) {
  for (col in 1:NCOL(X)) {
    maxcol <- max(X[, col])
    mincol <- min(X[, col])
    if(maxcol!= mincol){
      X[, col] <- (X[, col] - mincol) / (maxcol - mincol)
    }
  }
  X
}
```

What are the ways of constructing an anomaly detection ensemble? Broadly, anomaly detection ensembles can be categorised into 3 camps.

1. Feature bagging
2. Subsampling
3. Using combination functions


## Feature bagging
Feature bagging uses different attribute subsets to find anomalies. In a dataset, observations are denoted by rows and attributes are denoted by columns. Feature bagging considers different column subsets. That is, multiple copies of the same dataset each having a slightly different set of columns is considered. For each dataset copy, we find anomalies using a single anomaly detection method.  Then the anomaly scores are averaged to compute the ensemble score. 

Let us try this with the letter dataset from the [ODDS repository](http://odds.cs.stonybrook.edu/). We first read the dataset and normalize it so that each column has values within 0 and 1. 

```{r featurebagging}
datori <- readMat("letter.mat")
Xori <- datori$X
Xori <- unitize(Xori)
```